### 2023-05-15 tex/linear.tex
[ writing ] about PD modal logic

The techniques of \cref{sec:simple} were developed to handle idealised core
calculi, like the simply typed $\lambda$-calculus and its many variants.
Such core calculi are fairly close to some real intermediate languages,
\todo{cite} though many intermediate languages also have polymorphism in their
type systems, which I have not discussed, but has been considered by much prior
work\todo{cite}.

%Linear logic was introduced by \citet{girard87linear} as a tool of proof theory.
%Its distinguishing feature is that hypotheses, by default, must be used exactly
%once.
%I devote a chapter to introducing linear logic because it is better understood
%than any individual or general semiring-annotated calculi, and is thus an
%important point of comparison.
%Linear logic will be an instance of the semiring-annotated systems we study
%later.

>>>>

The techniques of \cref{sec:simple} were developed to handle idealised core
calculi, like the simply typed $\lambda$-calculus and its many variants.
Such core calculi are fairly close to some real intermediate languages,
\todo{cite} though many intermediate languages also have polymorphism in their
type systems, which I have not discussed, but has been considered by much prior
work~\citep{POPLmark}.

Another potential feature --- required for many desired semantics but only
relatively recently appearing in programming languages --- that is not handled
by the techniques of \cref{sec:simple} is restrictions on how variables can be
used.
In the simple type systems covered in \cref{sec:simple}, the only restriction on
the use of a variable is that it is used as an expression of the appropriate
type.
However, we could also consider having somehow ``guarded'' subexpressions in
which some variables bound outside cannot occur, or systems in which the use of
a variable in one part of an expression precludes its use in another part of
that expression (the variable having been ``used up'').
I will introduce and motivate some such systems in this chapter, and argue that
they are impossible to capture directly in the framework presented in
\cref{sec:gen-syn}.

Languages and systems implementing features inspired by linear logic include
Rust~\citep{MK14,Rust}, ATS~\citep{Xi04,ZX05}, and various implementations of
session types~\citep{HLVCCDMPRTVTZ16}.
Additionally, forms of linearity have been used in theoretical work, for example
to bound computational complexity~\citep{GSS92,Hofmann03} or maintain
differentiability~\citep{ER03,Ehrhard18}.
As for modality, \citet{IMO20} show that common data-flow analyses for
imperative languages can be recast as modal type systems, though they have
generally been developed in an ad hoc fashion.
Such a recasting may be useful to understand the metatheoretic properties of
programs passing such static analyses.
More user-facing implementations of modal type systems include the system for
stack allocation in OCaml~\citep{DW22} and the approach to erasure found in
Agda and Idris\todo{cite}.

In this chapter\ldots\todo{Chapter map}

\section{Intuitionistic S4 modal logic}\label{sec:modal}
\input{modal.tex}

### 2023-05-15 tex/rensub.tex
[ writing ] about PD modal logic

I have made few unforced design decisions in this chapter, so there are few
aspects I can see future work improving upon.\todo{Which?}
However, future work may want to extend the work of this chapter, in which case
there are some unanswered questions.
Principal among these, in my mind, is dealing with equivalence/equality of
environments.
We want to talk about equality of environments for two related purposes.
The most immediate is that we want to develop the equational theory of renaming
and substitution --- for example, when we use \cref{thm:env-comp} to compose
substitutions, we expect that composition to be associative and unital (with
respect to \cref{thm:env-id}).
These equations of substitutions should yield equations on terms in which such
substitutions have been applied.
Slightly more abstractly, I would like to develop a theory of
\emph{quantitative multicategories}, in which multimorphisms have in their
domain a list of objects paired with usage annotations.
I would hope for $\name$ types and terms to give an example quantitative
multicategory, analogously to how Lambek calculus gives an example of an
ordinary multicategory and the simply typed $\lambda$-calculus gives an example
of a Cartesian multicategory.

>>>>

Future work may want to extend the work of this chapter, in which case
there are some unanswered questions.
Principal among these, in my mind, is dealing with equivalence/equality of
environments.
We want to talk about equality of environments for two related purposes.
The most immediate is that we want to develop the equational theory of renaming
and substitution --- for example, when we use \cref{thm:env-comp} to compose
substitutions, we expect that composition to be associative and unital (with
respect to \cref{thm:env-id}).
These equations of substitutions should yield equations on terms in which such
substitutions have been applied.
Slightly more abstractly, I would like to develop a theory of
\emph{quantitative multicategories}, in which multimorphisms have in their
domain a list of objects paired with usage annotations.
I would hope for $\name$ types and terms to give an example quantitative
multicategory, analogously to how Lambek calculus gives an example of an
ordinary multicategory and the simply typed $\lambda$-calculus gives an example
of a Cartesian multicategory.
